Tempus: Probabilistic Network Latency Verifier

### 2022-11-17 paper/main.tex
[Paper] Updated chapter 3

\subsubsection{Numerical Convolution}
One problem with this algorithm, is that not every convolution between distributions have a 
closed-form solution and could be done analytically.
Therefore, we need to leverage a numerical convolution method in order to convolve two arbitrary
distributions.

%TODO: explain the convolution, weighting, DIRECT, changing order of DIRECT
Expand points on 
\begin{itemize}
    \item DIRECT 
    \item orderings DIRECT
\end{itemize}

\subsubsection{Equivalence Class Consolidation}
The first is what we call \textbf{Equivalence Class Consolidation}. 
The idea is, while each hot-edges events has the same convergent paths, it is not guaranteed to be 
unique. 
Two or more of these events can have the same convergent paths.
Therefore, we could merge these two events and compute the convolution for a given path once.

\subsubsection{Path Memoization}

\subsubsection{Convolution Grouping}

>>>>

\subsubsection{Numeric Convolution}
One potential problem with a convolution-based technique is not every distribution pair can be convolved 
analytically.
A closed-form solution of a convolution is usually only available for two distributions of the same type.
Therefore, in order to convolve two arbitrary distributions, we need to leverage a numerical convolution 
method.

We leverage an existing numerical convolution algorithm, DIRECT, to fulfill this role.
%TODO: describe DIRECT in one sentence
We choose this method due to their bounded error property: DIRECT guarantee that the computed 
distribution and the correct theoretical distribution has a KL-divergence below a certain bound.
DIRECT has also been implemented in R's popular bayesmeta package.
%TODO: cite and change font?

One subtle detail about DIRECT is while convolution is defined to be a commutative operation, and 
DIRECT also achieves this property, the order of operation matters for the algorithm's runtime 
performance.
In particular, if we had a chain of DIRECT convolutions, the result of a convolution should not be set 
as the second input distribution in the later convolutions to avoid performance penalty.
This is caused by a nested loop from the following two mechanisms:
\begin{itemize}
    \item DIRECT works by computing a list of support values by doing many PDF queries of the second input
        distribution
    \item Computing the PDF of a DIRECT distribution (the returned distribution of a DIRECT 
        algorithm) itself involves iterating over the current support values
\end{itemize}

As an ilustration, if we had 3 random variables $A$, $B$, and  $C$, and we want 
to convolve them all together, it would be faster to compute $direct(direct(A, B), C)$ instead of 
$direct(C, direct(A, B))$.

While this problem is trivial to solve (swap the input parameter if the latter is detected as a 
DIRECT distribution), it does limit the type of optimization that we could have in the chain convolution 
process.

\subsubsection{Equivalence Class Consolidation}
To minimize the amount of temporal verification overhead, we want to find some symmetry in the 
existing exploration algorithm so that we could reduce the exploration space even further.

In the original NetDice exploration algorithm, multiple network states were merged into an equivalence 
class based on its cold edges (edges whose failure won't change the convergent path(s)).
In other words, network states within the same equivalence class will have the same convergent path(s).

However, we observe that each of those equivalence classes are not \textit{unique}: while network states within 
the same equivalence class shares a convergent path(s), \textbf{multiple equivalence classes could also 
share the exact same convergent path(s)}.
Since we define latency distribution to marginalize over all factors other than the path, we could 
effectively \textit{consolidate} these equivalence classes into one big equivalence class and do temporal 
verification once.

We implemented this idea by doing memoization on the temporal probability of a given convergent path(s).
We will only do temporal probability computation once, when we first iterate over an equivalence class that 
has a certain convergent path(s), and we cached the result should another equivalence class with the same 
convergent path(s) emerged.

\subsubsection{Path Memoization}
After consolidating many equivalence classes that shares the same convergent path(s), we are left with fewer, 
bigger, but unique equivalence classes.
As significant as it is, we could reduce the computation cost further by drawing connections between these 
unique equivalence classes.

In a network with a load-balancing protocol, the convergent state of the data plane might forward a packet 
through multiple possible paths.
In the context of our framework, we say that an equivalence class might have more than one convergent paths.
Nevertheless, we note that two equivalence classes with different convergent paths might not be two 
independent subset.
In other words, \textbf{multiple equivalence classes could share a subset of individual path}.

%TODO: example AB and A, only need to do reweighting

Since the way we compute convergent path(s) temporal probability is by calculating the weighted average of 
individual path temporal probability, we could reduce the amount of convolution we will need to do by doing further memoization 
on the temporal probability of a given path.
If we were to compute the temporal probability of an equivalence class with previously cached individual 
path, we only need to calculate the weights that corresponds to the load balancing protocol without 
doing any convolution.

\subsubsection{Convolution Grouping}
Aside from optimizing the \textit{amount} of paths we need to do convolution on, we could also do some 
optimization on the \textit{convolution operation itself}.

While the use of numerical technique to do convolution enabled us to use arbitrary probability distribution 
in our framework, it comes with two noticeable disadvantages.

The first is \textit{error}: numerical method is fundamentally an approximation technique that always 
comes with some error. 
While DIRECT algorithm comes with an error-bound guarantee, it is not zero and could expand if we do 
chain convolution.

The second is \textit{runtime performance}: even with the input ordering constraints we laid out on the 
previous subsection, DIRECT is still generally slower than analytical convolution.
This is expected, since a closed form solution of a convolution tipically boils the operation down into 
a few arithmetic calculation, rather than complex operation like integration.

Due to these factors, we feel the need to \textbf{prioritize analytical convolution} in the calculation of 
path latency distribution.
This is done by \textit{grouping} subsets of the path components with the same probability distribution 
type (e.g. exponential, normal, etc.). 
We then analytically convolve the distribution within that subset and only numerically convolve the 
returned distribution, which will be fewer in quantity, thus enabling smaller error and faster performance.
